Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2016-11-15 draft.tex
.

To overcome this contradiction, we introduce an idealized picture of diffusive homeostasis, where diffusion across the tissue is instantaneous. This simplification results in a single level of NO-concentration for all neurons, only being modified over time by the sum of all neurons' NO-syntheses and the decay term $-\lambda NO$. To account for the spread of NO, $nNOS(t)$ must be divided by the area $L^2$ of the tissue. The idealized version of \eqref{NO_dyn} then reads
\begin{equation}
\dot{NO}(t)=-\lambda NO + \frac{\sum_{i} nNOS^i(t)}{L^2}\;.
\label{NO_dyn_approx}
\end{equation}
In this form, it becomes clear that we can now describe the \emph{total} influx of NO (the sum in \eqref{NO_dyn_approx}) as the sum of a mean and a random fluctuation: While spiking events on individual cells remain sufficiently separated to justify equation \eqref{nNOS_single_sol_approx2}, the sum of all spike events from the excitatory population results in an effective poisson process with a mean rate of $3 Hz \cdot N_{exc.} = \mathrm{1200\,Hz}$, given the assumption that cells are spiking weakly correlated. Furthermore, we argue that a total rate of $\mathrm{1200\,Hz}$ is sufficiently large to express this rate as a continuous function of time, $r_{total} = N_{exc.}\cdot r_{exc.pop.}(t)$, at least on the time scale of the oscillations to be studied (see  Figure \ref{rate_th_close}). Thus, equation \eqref{NO_dyn_approx} becomes
\begin{equation}
\dot{NO}(t)=-\lambda NO + \frac{\mu_{nNOS}(N_{exc.}r_{exc.pop.}(t)) + \sigma_{nNOS}(N_{exc.}r_{exc.pop.}(t)) \xi (t)}{L^2}\;.
\label{NO_dyn_approx2}
\end{equation}
As a next step, we attempted to simplify the dynamics of threshold adaption and its effect on the population rate. Since equation \eqref{NO_dyn_approx2} contains a single variable $r_{exc.pop.}(t)$ for the mean excitatory activity, we attempted to find an appropriate description based solely on

@@

To complete the expression, one has to find an appropriate description of the $r(\theta)$-relation. As a most simple approximation, we applied a linear regression to the data of the average threshold and the corresponding population activity, see Figure \ref{thresh_r_linfit}. The constant $C$ in equation \eqref{simple_NO_dyn} is then proportional to the slope of the linear fit. In addition, $C$ is proportional to the amount of NO released per spike, proportional to the number of contributing neurons and antiproportional to the total area of the tissue. We can solve and integrate equations \eqref{Ca_dyn} and \eqref{nNOS_dyn} for a single spike, resulting in a total of $\tau_{Ca^{2+}}\cdot ln(1+{Ca^{2+}_{spike}}^3) / 3$. Therefore, $C=N_{neur}\cdot \tau_{Ca^{2+}}\cdot ln(1+{Ca^{2+}_{spike}}^3) / (3\cdot L^2)$ and $C\cdot r(\theta)=\theta \cdot \alpha \cdot N_{neur} \cdot \tau_{Ca^{2+}}\cdot ln(1+{Ca^{2+}_{spike}}^3) / (3\cdot L^2)$, where $\alpha$ is the slope of the linear fit and $L^2$ the area of the tissue.
\begin{figure}
\includegraphics[width=\textwidth]{../../plots/diff_hom/thresh_r_linfit.eps}
\caption{Linear fit of the relation between the firing threshold (averaged over population) and the population activity. $R^2=0.99996$.}
\label{thresh_r_linfit}
\end{figure}
\begin{figure}
\includegraphics[width=\textwidth]{../../plots/diff_hom/lin_model_osc.eps}
\caption{Simulation results for equations \eqref{simple_NO_dyn} and \eqref{simple_threshold_dyn}.}
\label{lin_mod_osc}
\end{figure}
Figure \ref{lin_mod_osc} shows the results achieved from the simulation. We observe that noise is capable of maintaining oscillations in an otherwise stable linear system.

>>>>

To overcome this contradiction, we introduce an idealized picture of diffusive homeostasis, where diffusion across the tissue is instantaneous. This simplification results in a single level of NO-concentration for all neurons, only being modified over time by the sum of all neurons' NO-syntheses and the decay term $-\lambda NO$. To account for the spread of NO, $nNOS(t)$ must be divided by the area $L^2$ of the tissue. The idealized version of \eqref{NO_dyn} then reads
\begin{equation}
\dot{NO}(t)=-\lambda NO + \frac{\sum_{i} nNOS^i(t)}{L^2}\;.
\label{NO_dyn_approx}
\end{equation}
In this form, it becomes clear that we can now describe the \emph{total} influx of NO (the sum in \eqref{NO_dyn_approx}) as the sum of a mean and a random fluctuation: While spiking events on individual cells remain sufficiently separated to justify equation \eqref{nNOS_single_sol_approx2}, the sum of all spike events from the excitatory population results in an effective poisson process with a mean rate of $3 Hz \cdot N_{exc.} = \mathrm{1200\,Hz}$, given the assumption that cells are spiking weakly correlated. Furthermore, we argue that a total rate of $\mathrm{1200\,Hz}$ is sufficiently large to express this rate as a continuous function of time, $r_{total} = N_{exc.}\cdot r_{exc.pop.}(t)$, at least on the time scale of the oscillations to be studied (see  Figure \ref{rate_th_close}). Thus, equation \eqref{NO_dyn_approx} becomes
\begin{equation}
\dot{NO}(t)=-\lambda NO + \frac{\mu_{nNOS}(N_{exc.}r_{exc.pop.}(t)) + \sigma_{nNOS}(N_{exc.}r_{exc.pop.}(t)) \xi (t)}{L^2}\;.
\label{NO_dyn_approx2}
\end{equation}
As a next step, we attempted to simplify the dynamics of threshold adaption and its effect on the population rate. Since equation \eqref{NO_dyn_approx2} contains a single variable $r_{exc.pop.}(t)$ for the mean excitatory activity, we wanted to find an appropriate description containing the mean threshold $V_{t,pop.}(t) = <V_t(t)>$. As stated earlier, Figure \ref{rate_th_close} suggests an immediate and approximately linear relation between excitatory population rate and the mean of thresholds within the excitatory population. Figure \ref{thresh_r_linfit} shows a linear fit (being of the form $\alpha x + \beta$) of these two quantities plotted against each other. Apart from rarely appearing high values of $V_{t,pop.}$, we found a good fit to the linear model and concluded $r_{exc.pop.}(t) \approx r_{IP} + \alpha(V_{t,pop.}(t)-V_{t,0})$, where $r_{IP}$ is given in table \ref{Params_IP} as the mean excitatory target firing rate and $V_{t,0} \equiv (r_{IP}-\beta)/\alpha$ is the mean threshold corresponding to the target rate.
\begin{figure}
\includegraphics[width=\textwidth]{../../plots/Spike_Stats/linfit_th_rate.png}
\caption{Linear fit of excitatory population activity and mean excitatory threshold. Slope $\alpha$ of the fit: $\mathrm{-0.772\, Hz/mV}$. Offset $\beta$: $\mathrm{-41.01\, Hz}$. Mean squared error: $\mathrm{R^2 = 0.69}$.}
\label{thresh_r_linfit}
\end{figure}

Combining these results in a set of equations, we find
\begin{align}
\begin{split}
\dot{NO}(t) &= - \lambda NO + \frac{ N_{exc.} {Ca^{2+}_{spike}}^3 \tau_{Ca^{2+}}ln(2)}{3 L^2}[\alpha(V_{t,pop.}(t)-V_{t,0}) + r_{IP}] \\
&+ \frac{{Ca^{2+}_{spike}}^3 {\tau_{Ca^{2+}}} ln(2)}{3L^2}\sqrt{\frac{N_{exc.}}{2 \tau_{nNOS}}}\sqrt{\alpha(V_{t,pop.}(t)-V_{t,0}) + r_{IP}}\xi (t)
\end{split}\label{2d-sys-homeostasis1}\\
\dot{V_{t,pop.}}(t) &= \frac{NO-NO_0}{NO_0 \tau_{V_t}}\;. \label{2d-sys-homeostasis2}
\end{align}
Through the coordinate transformation $n = NO-NO_0 \; , \theta = V_{t,pop.}(t)-V_{t,0}$, we can a simplify the dynamical system:
\begin{align}
\begin{split}
\dot{n} &= -\lambda n + \frac{\alpha N_{exc.} {Ca^{2+}_{spike}}^3 \tau_{Ca^{2+}}ln(2)}{3 L^2}\theta \\
&+ \frac{{Ca^{2+}_{spike}}^3 {\tau_{Ca^{2+}}} ln(2)}{3L^2}\sqrt{\frac{N_{exc.}}{2 \tau_{nNOS}}}\sqrt{\alpha\theta + r_{IP}}\xi (t)
\end{split}\label{2d-sys-homeostasis-coord_transf1}\\
\dot{\theta} &= \frac{1}{NO_0 \tau_{V_t}} n \label{2d-sys-homeostasis-coord_transf2}
\end{align}

@@

To complete the expression, one has to find an appropriate description of the $r(\theta)$-relation. As a most simple approximation, we applied a linear regression to the data of the average threshold and the corresponding population activity, see Figure \ref{thresh_r_linfit}. The constant $C$ in equation \eqref{simple_NO_dyn} is then proportional to the slope of the linear fit. In addition, $C$ is proportional to the amount of NO released per spike, proportional to the number of contributing neurons and antiproportional to the total area of the tissue. We can solve and integrate equations \eqref{Ca_dyn} and \eqref{nNOS_dyn} for a single spike, resulting in a total of $\tau_{Ca^{2+}}\cdot ln(1+{Ca^{2+}_{spike}}^3) / 3$. Therefore, $C=N_{neur}\cdot \tau_{Ca^{2+}}\cdot ln(1+{Ca^{2+}_{spike}}^3) / (3\cdot L^2)$ and $C\cdot r(\theta)=\theta \cdot \alpha \cdot N_{neur} \cdot \tau_{Ca^{2+}}\cdot ln(1+{Ca^{2+}_{spike}}^3) / (3\cdot L^2)$, where $\alpha$ is the slope of the linear fit and $L^2$ the area of the tissue.
\begin{figure}
\includegraphics[width=\textwidth]{../../plots/diff_hom/thresh_r_linfit.eps}
\caption{Linear fit of the relation between the firing threshold (averaged over population) and the population activity. $R^2=0.99996$.}

\end{figure}
\begin{figure}
\includegraphics[width=\textwidth]{../../plots/diff_hom/lin_model_osc.eps}
\caption{Simulation results for equations \eqref{simple_NO_dyn} and \eqref{simple_threshold_dyn}.}
\label{lin_mod_osc}
\end{figure}
Figure \ref{lin_mod_osc} shows the results achieved from the simulation. We observe that noise is capable of maintaining oscillations in an otherwise stable linear system.
